Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-20 paper/main.tex

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions. What result might we expect to see if this is the case? To
gain an intuition for this possibility, consider the expected outcome if we
carried out these same analyses using a simple proportion-correct measure in
lieu of our knowledge estimates. Suppose a participant correctly answered $n$
out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n - 1}{q - 1}$. If we hold out a single
\textit{incorrectly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n}{q - 1}$. In this way, the proportion of
correctly answered remaining questions is always \textit{lower} when the
held-out question was answered correctly than when it was answered incorrectly.
Because our knowledge estimates are computed as a weighted version of this same
proportion-correct score (where each held-in question's weight reflects its
embedding-space distance from the held-out question; see Eqn.~\ref{eqn:prop}),
if these weights were uninformative (e.g., randomly distributed), then we
should expect to see this same inverse relationship between estimated knowledge
and performance, on average. On the other hand, if the spatial relationships
among the quiz questions' embeddings \textit{are} predictive of participants'
knowledge about the questions' content, then we would expect \textit{higher}
estimated knowledge for held-out correctly (versus incorrectly) answered
questions.

@@

>>>>

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions. What result might we expect to see if this is the case? To
gain an intuition for this possibility, consider the expected outcome if we
carried out these same analyses using a simple proportion-correct measure in
lieu of our knowledge estimates. Suppose a participant correctly answered $n$
out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n - 1}{q - 1}$. Whereas if we hold out a single
\textit{incorrectly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n}{q - 1}$. In this way, the proportion of
correctly answered remaining questions is always \textit{lower} when the
held-out question was answered correctly than when it was answered incorrectly.
Because our knowledge estimates are computed as a weighted version of this same
proportion-correct score (where each held-in question's weight reflects its
embedding-space distance from the held-out question; see Eqn.~\ref{eqn:prop}),
if these weights are uninformative (e.g., randomly distributed), then we
should expect to see this same inverse relationship between estimated knowledge
and performance, on average. On the other hand, if the spatial relationships
among the quiz questions' embeddings \textit{are} predictive of participants'
knowledge about the questions' content, then we would expect \textit{higher}
estimated knowledge for held-out correctly (versus incorrectly) answered
questions.

@@

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed whether estimates of participants' knowledge at the embedding coordinates of individual quiz questions could reliably predict their ability to correctly answer those questions.
In essence, we treated each question a given participant answered on a given quiz as a ``lecture'' consisting of a single timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's knowledge for its embedding coordinate based on their performance on all \textit{other} questions they answered on that same quiz (``All questions''; Fig.~\ref{fig:predictions}, top row).
We also computed analogous knowledge estimates for each lecture-related question based on all other questions (answered by the same participant, on the same quiz) about the same lecture (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows) and based on all questions about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of knowledge estimates for all questions answered by all participants on the given quiz.
Our models were implemented in \texttt{R} using the \texttt{lme4} package \citep{BateEtal15}.

%In each versions of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of

%In the set of analyses reported in Figure~\ref{fig:predictions}, we used generalized linear mixed models (GLMMs) to analyze whether estimates of participants' knowledge at the embedding coordinates of individual held-out questions could reliably predict their ability to answer those questions correctly.

$$
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
$$

note that qID is implicitly nested within levels of lecture

$$
\begin{aligned}
  \operatorname{accuracy}_{i}  &\sim \operatorname{Binomial}(n = 1, \operatorname{prob}_{\operatorname{accuracy} = \operatorname{correct}} = \widehat{P}) \\
    \log\left[\frac{\hat{P}}{1 - \hat{P}} \right] &=\alpha_{j[i],k[i]} \\
\left(
  \begin{array}{c}
    \begin{aligned}
      &\alpha_{j} \\
      &\gamma_{1j}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c}
    \begin{aligned}
      &\gamma_{0}^{\alpha} + \gamma_{1k[i]}^{\alpha}(\operatorname{knowledge}) \\
      &\mu_{\gamma_{1j}}
    \end{aligned}
  \end{array}
\right)
,
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{j}} & \rho_{\alpha_{j}\gamma_{1j}} \\
     \rho_{\gamma_{1j}\alpha_{j}} & \sigma^2_{\gamma_{1j}}
  \end{array}
\right)
 \right)
    \text{, for participant j = 1,} \dots \text{,J} \\
\left(
  \begin{array}{c}
    \begin{aligned}
      &\alpha_{k} \\
      &\gamma_{1k}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c}
    \begin{aligned}
      &\mu_{\alpha_{k}} \\
      &\mu_{\gamma_{1k}}
    \end{aligned}
  \end{array}
\right)
,
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{k}} & \rho_{\alpha_{k}\gamma_{1k}} \\
     \rho_{\gamma_{1k}\alpha_{k}} & \sigma^2_{\gamma_{1k}}
  \end{array}
\right)
 \right)
    \text{, for question k = 1,} \dots \text{,K}
\end{aligned}
$$
